Taking Dyads Seriously

### 2018-02-09 paper/4_app.tex
even more changes to ap section

@@

In order to showcase the insights the AME approach permits scholars to
bring into their analysis, we illustrate the application of AME on
five recent studies from the broad field of international relations
and international political economy that utilize relational data
\citep{reiter:stam:2003, mcdonald:2004,  rose:2004, weeks:2012,
  gibler:2017}.  These articles demonstrate the broad range of
application for the AME framework, which include testing the effect of
both dyadic and monadic variables on varied conflict behaviors and
trade.  We chose to demonstrate the capabilities of AME with reference
to existing studies in order to highlight several features of the
method. Firstly, the results of AME estimation are interpretable
alongside GLM results. Secondly, although the data requirements of AME
are, admittedly, more intensive than needed for GLMs, this rich data
is increasingly coming online in both new and existing data. Thirdly,
through AME, we can examine the extent to which interdependencies
within typical dyadic data alter previous conclusions about
international relations from recent empirical studies using a
framework that a priori assumes conditional independence.

These studies were also selected based on how recent they were and
whether they had more then 100 citations.\footnote{Note that we chose
  papers with at least 100 citations as an indicator of influence
  within the discipline. Of course, by the criteria of influence many
  other papers could have been considered for the
  replication. However, as described in the following sections, the
  AME model has specific data requirements, which limited the scope of
  potential papers.}  Each of these pieces, published in prominent
journals well-known in their respective literatures, posited a theory
in which interdependencies are consequential. However, reflecting the
dominant approach in the literature, each of the authors tested their
hypothesis by employing some form of a general linearized
regression.

In the following section, we present two sets of results for each
study, one that follows the original study in our re-estimation of the
original models and another that tests the central hypothesis within
an AME implementation. We demonstrate that AME allows scholars the
ability to test their hypotheses of interest, while directly
accounting for both theorized and observed interdependencies in the
data. Directly accounting for interdependences can change the outcome
of regression models, both in substance and in estimates. In part, this outcome
follows from AME's ability to precisely capture higher-order
interdependencies, whereas GLM approaches implicitly ignore dyadic
interdependencies. As this is a widely-known limitation of GLM
approaches, scholars often attempt to account for clustering of
observations by including additional variables and adjusting the
standard errors of the resulting estimates. At best, this method
introduces noise and imprecision into results, and at worst can
produce misleading outcomes.

@@

An important takeaway here is that as a result of the incompatibility
between GLM models and the underlying processes that shape
international events, many scholars are forced to make knowledge claims
based on the statistical significance of a small set of covariates, or
the differences between these covariates. These differences may change
dramatically when interdependencies are taken into account directly,
as they do in the case of this study. Beyond just comparing parameter
estimates, we examine how well each approach can represent the data
generating process using an out-of-sample cross validation
strategy. Specifically, for each study, we randomly divide the data
into $k=30$ sets, letting $s_{ij,t}$ be the set to which pair $ij,t$
is assigned.

@@

\citet{mcdonald:2004} studies whether trade promotes peace between nations. The link between conflict and trade is perplexing, with many persistent, yet competing explanations. \citet[p. 547]{mcdonald:2004} includes the argument that interdependence between states ``makes conflict less likely because of its efficiency over conquest in acquiring resources\ldots''. Accordingly, his primary contribution is to provide evidence challenging the generalized linkage between peace and trade and to offer a new measurement of the key independent variable, trade. \citet{mcdonald:2004} refines the trade variable, arguing that \textit{free} trade, rather than trade alone, reduces the likelihood of conflict between states. His key hypothesis is that greater levels of protection increase the probability of interstate conflict, an argument that builds on the work of classic liberalism and connects free trade to the power of domestic audiences. \citet{mcdonald:2004} measures free trade in two ways. The first measurement reflects the notion that larger protected sectors generate greater societal pressures resulting in pockets of support for war. This protection variable measures the proportion of customs revenue divided by total imports in the state that possesses the greater such ratio in each dyad. This measure thus captures the score of the state in the dyad that possesses higher barriers to trade. \citet[p. 560]{mcdonald:2004} also includes a measure of economic integration  calculated as ``the lower proportion of total dyadic trade (imports plus exports) divided by state $i$'s GDP or total dyadic trade divided by state $j$'s GDP''. The binary, dependent variable is the onset of a new militarized interstate dispute within a given dyad. \citet{mcdonald:2004} employs logistic regression to examine the putative statistical significance of these variables. The models include splines to correct for temporal dependence, and robust standard errors clustered on each dyad.

@@

\citet{weeks:2012} examines the influence of domestic institutions on the initiation of military conflicts by autocratic leaders.  She argues that in some circumstances autocrats are held accountable for their foreign policy decisions. She adds the nuance that autocratic audiences are not homogeneous. When the autocratic regime is nonmilitary, domestic audiences do not favor military actions, but in military autocracies this is not the case. Further she argues that in personalistic regimes without a military or civilian domestic audience, the leaders tend to be more likely to employ military force in their foreign policy.  To study this question, she uses a dyadic design in which the dependent variable is ``whether country A in a directed dyad initiated military conflict against country B during year t'' (page 337).  One major innovation in her study resides in the nuanced way in which she conceptualizes and codes regime type into four types: a) Machine, b) Junta, c) Boss, and d) Strongmen. She also includes a variety of putative control variables focusing on capabilities for both sides of the dyad, alliances, geography, trade dependence, regime instability, and the regime type of ``side B.''  She uses a logistic regression, but follows \citet{beck:etal:1998} and includes splines to capture temporal covariation in the dependent variable along with fixed, unit effects. The analysis is done for dyads, but is considered to be from the perspective of the actor that initiated the dispute. The basic finding is that a) juntas, bosses, and strongmen are more likely to initiate conflict than machines (and maybe democracies) and that b) machines are no more belligerent than democracies.  These insights are mainly emphasized in the paper by the parameter estimates depicted in Tables 1 and 2 (pages 339--340) from the paper. She makes the argument that ignoring important nuances between different types of autocracies hinders our understanding of the initiation of military conflict by autocracies.

@@

The replications we have undertaken are all from articles over the past fifteen years.  A more recent example is \citet{gibler:2017} which examines the onset of militarized disputes using capabilities, joint democracy, alliances, and power parity in a undirected dyadic study using logistic regression and dyad clustered standard errors. In addition, Gibler shows that the long-standing relationship between the relative parity of capabilities and initiation of international conflict is almost completely mediated by the initial conditions for the members of the dyad when they joined the international system as sovereign members. This finding calls into question many IR theories about the role of balance in terms of generating international conflict \citep{organski:1958}.

@@

>>>>

\documentclass{article}[12pt]
\usepackage{setspace}
\usepackage{listings}
\doublespacing
\usepackage[letterpaper, portrait, margin=1in]{geometry}
\setlength{\parskip}{.5em}
\setlength{\parindent}{0cm}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{framed}
\usepackage[sort,round]{natbib}
\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{hyperref}

\begin{document}

@@

In order to showcase the insights the AME approach permits scholars to
bring into their analysis, we apply AME to five recent studies from the broad field of international relations
and international political economy~\citep{reiter:stam:2003, mcdonald:2004,  rose:2004, weeks:2012,
  gibler:2017}. Each of these studies are representative of broader trends in the field in that they use
relational data of state interactions and propose both dyadic,
monadic, and structural explanations for behavior of actors in the system.

We chose to demonstrate the capabilities of AME with reference
to existing studies in order to highlight several features of the
AME approach. Firstly, the results of AME estimation are interpretable
alongside GLM results but have the additional benefit of being able to
quantify higher-order features of interest to scsholars, such as
``sender'' and ``receiver'' effects. Secondly, although the data
requirements of AME are, admittedly, more intensive than needed for GLMs, ever richer data
is increasingly prevalent in both new and existing data. Thirdly,
through AME, we can examine the extent to which interdependencies
within typical dyadic data alter previous conclusions about
international relations from recent empirical studies using a
framework that a priori assumes conditional independence.

The five studies replicated below were selected based on how recent they were and
whether they had more then 100 citations.\footnote{Note that we chose
  papers with at least 100 citations as an indicator of influence
  within the discipline. Of course, by the criteria of influence many
  other papers could have been considered for the
  replication. However, as described in the following sections, the
  AME model has specific data requirements, which limited the scope of
  potential papers.}  Each of these pieces, published in prominent
journals well-known in their respective literatures, posited a theory
in which interdependencies are consequential. However, reflecting the
dominant approach in the literature, each of the authors tested their
hypothesis by employing some form of a general linearized
regression.

In the following section, we present two sets of results for each
study, one that replicates the original study and another that tests
the central hypothesis within an AME implementation. We demonstrate
that AME allows scholars the ability to test their hypotheses of
interest, while directly accounting for both theorized and observed interdependencies in the
data. As accounting for interdependences can change the outcome
of regression models, both in substance and in estimate, we also summarize the
predictive performance of the GLM and AME implementations. This
predictive view allows us to better compare the models across
specifications. Finally, for several of the studies, we report the
AME-derived sender and receiver effects in order to highlight
additional value of the AME framework.

@@

An important takeaway here is that as a result of the incompatibility
between GLM models and the underlying processes that shape
international events, many scholars are forced to make knowledge claims
based on the statistical significance of a small set of covariates, or
the differences between these covariates.  These differences may
change dramatically when interdependencies are taken into account
directly, as they do in the case of this study. %MJF: Keeping this here for now, but it reads more like a conclusion
This outcome follows from AME's ability to precisely capture higher-order
interdependencies, whereas GLM approaches implicitly ignore dyadic
interdependencies. As this is a widely-known limitation of GLM
approaches, scholars often attempt to account for clustering of
observations by including additional variables and adjusting the
standard errors of the resulting estimates. At best, this method
introduces noise and imprecision into results, and at worst can
produce misleading outcomes.   Beyond just comparing parameter
estimates, we examine how well each approach can represent the data
generating process using an out-of-sample cross validation
strategy. Specifically, for each study, we randomly divide the data
into $k=30$ sets, letting $s_{ij,t}$ be the set to which pair $ij,t$
is assigned.

@@

\citet{mcdonald:2004} studied whether trade promotes peace between
nations.  \citet[p. 547]{mcdonald:2004} introduced the argument that
free trade between states ``makes conflict less likely
because of its efficiency over conquest in acquiring
resources\ldots''. Accordingly, he provided evidence challenging the
generalized linkage between peace and trade and refined the
measurement of the standard ``trade'' variable, arguing that
\textit{free} trade, rather than trade alone, reduces the likelihood
of conflict between states. His key hypothesis is that greater levels
of protection increase the probability of interstate conflict, an
argument that builds on the work of classic liberalism and connects
free trade to the power of domestic audiences. \citet{mcdonald:2004}
measured free trade in two ways. The first, the ``protection''
variable measures the proportion of customs revenue divided by total
imports in the state that possesses the greater such ratio in each
dyad. reflects the notion that larger protected sectors generate
greater societal pressures resulting in pockets of support for war.
This measure thus captures the score of the state in the dyad that
possesses higher barriers to trade. \citet[p. 560]{mcdonald:2004} also
includes a measure of economic integration  calculated as ``the lower
proportion of total dyadic trade (imports plus exports) divided by
state $i$'s GDP or total dyadic trade divided by state $j$'s
GDP''. The (binary) dependent variable is the onset of a new
militarized interstate dispute within a given dyad.
 \citet{mcdonald:2004} employs logistic regression to examine the
 putative statistical significance of these variables. The models
 include splines to correct for temporal dependence, and robust
 standard errors clustered on each dyad.

@@

\citet{weeks:2012} examines the influence of domestic institutions on
the initiation of military conflicts by autocratic leaders.  She
argues that in some circumstances autocrats are held accountable for
their foreign policy decisions. She adds the nuance that autocratic
audiences are not homogeneous. When the autocratic regime is
nonmilitary, domestic audiences do not favor military actions, but in
military autocracies this is not the case. Further she argues that in
personalistic regimes without a military or civilian domestic
audience, the leaders tend to be more likely to employ military force
in their foreign policy.  To study this question, she uses a dyadic
design in which the dependent variable is ``whether country A in a
directed dyad initiated military conflict against country B during
year t'' (page 337).  One major innovation in her study resides in the
nuanced way in which she conceptualizes and codes regime type into
four types: a) Machine, b) Junta, c) Boss, and d) Strongmen. She also
includes a variety of putative control variables focusing on
capabilities for both sides of the dyad, alliances, geography, trade
dependence, regime instability, and the regime type of ``side B.''
She uses a logistic regression, but follows \citet{beck:etal:1998} and
includes splines to capture temporal covariation in the dependent
variable along with fixed, unit effects. The analysis is done for
dyads, but is considered to be from the perspective of the actor that
initiated the dispute. The basic finding is that a) juntas, boss type,
and strongmen type regimes are more likely to initiate conflict than
machine-type regimes (and maybe democracies) and that b) machine-type
regimes are no more belligerent than democracies.  
These insights are mainly emphasized in the paper by the parameter
estimates depicted in Tables 1 and 2 (pages 339--340) from the
paper. She argued that ignoring important nuances between
different types of autocracies hinders our understanding of the
initiation of military conflict by autocracies.

@@

A more recent example is \citet{gibler:2017} which examines the onset of militarized disputes using capabilities, joint democracy, alliances, and power parity in a undirected dyadic study using logistic regression and dyad clustered standard errors. In addition, Gibler shows that the long-standing relationship between the relative parity of capabilities and initiation of international conflict is almost completely mediated by the initial conditions for the members of the dyad when they joined the international system as sovereign members. This finding calls into question many IR theories about the role of balance in terms of generating international conflict \citep{organski:1958}.

@@

\end{document}
